Using VariantSpark to learn population structure from genomic profiles

### 2015-10-27 doc/publications/Recomb15/recomb15.tex
neaten

To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
DL method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.

@@

These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.

\FloatBarrier
\clearpage
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).

The mean ARI for our \kMeans clustering models on this dataset is \(0.960 \pm 0.004 \). 
%DENIS: lower compared to what ?
Although slightly lower, this value is similar to the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). 
Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.

%DENIS: I don't understand this 
To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we previously clustered
 these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.

Given the perfect result when building
the logistic regression model on this small subset of data, and the similarity with the other models,
we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.

@@

\bibliographystyle{plain}
\marginpar{TODO you need to find a way of shortening the Locket paper author list}
\bibliography{recomb}

>>>>

To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.

@@

%These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
%For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
%\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
%One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
%{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
%Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
%these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
%optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.

\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).

The mean ARI for our \kMeans{} clustering models on this dataset is \(0.960 \pm 0.004 \). 
This value is slightly less than the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual are misclassified as CHB). 
Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.

Although we see a slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
We therefore proceed to apply our method to the entire 1000 Genomes Project phase 1 dataset, which we previously clustered with an ARI of 0.84.

@@

\bibliographystyle{my_plain}
\bibliography{recomb}
